Racial and Ethnic Differences in ETI

### 2023-08-30 Kamyron Jordan/Racial and Ethnic Differences in ETI/racial_and_ethnic_differences_in_eti.qmd
started using table1

```{r libraries}
#| include: false
library(Hmisc)
library(tidyverse)
library(multidplyr)
library(gtsummary)
library(lmerTest)
library(broom.mixed)
library(emmeans)
library(knitr)
```

```{r data cleaning}
#| include: false
# source("/Users/timvigers/GitHub/CF/Kamyron Jordan/Racial and Ethnic Differences in ETI/create_race_eti_analysis_dataset.R")
load("~/Documents/Work/Vigers/CF/Kamyron Jordan/Racial and Ethnic Differences in ETI/Data_Cleaned/analysis_dataset.RData")
# Make modulators ordered factors
encounter <- encounter %>%
  mutate(across(contains("vx"), ~ factor(.x, ordered = T)))
```

@@

```{r}
demo %>%
  select(race, Hispanicrace) %>%
  tbl_summary(
    by = Hispanicrace,
    statistic = list(all_continuous() ~ "{mean} ({sd})")
  ) %>%
  add_overall() %>%
  bold_labels() %>%
  as_kable_extra()
```

@@

```{r}
annual %>%
  select(ReviewYear, race, Hispanicrace) %>%
  tbl_summary(
    by = ReviewYear, missing_text = "Missing",
    statistic = list(all_continuous() ~ "{mean} ({sd})")
  ) %>%
  bold_labels() %>%
  as_kable_extra()
```

@@

```{r}
demo %>%
  select(race) %>%
  tbl_summary(statistic = list(all_continuous() ~ "{mean} ({sd})")) %>%
  bold_labels() %>%
  as_kable_extra()
```

@@

```{r}
demo %>%
  select(race, Gender, WasGenotyped, F508_num) %>%
  tbl_summary(
    by = race, missing_text = "Missing",
    statistic = list(all_continuous() ~ "{mean} ({sd})")
  ) %>%
  add_overall() %>%
  bold_labels() %>%
  as_kable_extra()
```

@@

```{r}
#| cache: true
encounter %>%
  group_by(eDWID, reviewyear) %>%
  summarise(across(c(race, Gender, WasGenotyped, F508_num, contains("insurance")), first),
    `Age at Last Visit` = last(na.omit(encounterage)),
    `Mean ppFEV` = mean(GLI_FEV1_pct_predicted, na.rm = T),
    `Weight at Last Visit` = last(na.omit(weight)),
    `Weight %ile at Last Visit` = last(na.omit(weightpercentile)),
    `BMI at Last Visit` = last(na.omit(bmivalue)),
    `BMI %ile at Last Visit` = last(na.omit(bmipercentile)),
    across(contains("Vx"), ~ max(.x)),
    .groups = "drop"
  ) %>%
  select(
    -eDWID, reviewyear, race, Gender, WasGenotyped, F508_num, contains("insurance"),
    `Age at Last Visit`, `Mean ppFEV`, `Weight at Last Visit`,
    `Weight %ile at Last Visit`, `BMI at Last Visit`,
    `BMI %ile at Last Visit`, contains("Vx")
  ) %>%
  rename(any_of(tidy_labels)) %>%
  tbl_summary(
    by = Year, missing_text = "Missing",
    statistic = list(all_continuous() ~ "{mean} ({sd})")
  ) %>%
  bold_labels() %>%
  as_kable_extra()
```

@@

```{r}
#| warning: false
encounter %>%
  group_by(eDWID, reviewyear) %>%
  summarise(across(c(GLI_FEV1_pct_predicted, GLI_FVC_pct_predicted), ~ mean(.x, na.rm = T)),
    .groups = "drop"
  ) %>%
  pivot_longer(c("GLI_FEV1_pct_predicted", "GLI_FVC_pct_predicted")) %>%
  ggplot(aes(x = reviewyear, y = value, color = factor(eDWID), group = eDWID)) +
  geom_line(alpha = 0.01) +
  theme_bw() +
  theme(legend.position = "none") +
  facet_wrap(~name)
```

```{r}
#| cache: true
# Most basic model of lung function over time by race
mod <- lmer(
  GLI_FEV1_pct_predicted ~ factor(reviewyear) * race + (1 | eDWID),
  data = encounter %>% group_by(eDWID, reviewyear) %>%
    summarise(across(c(GLI_FEV1_pct_predicted, GLI_FVC_pct_predicted), ~ mean(.x, na.rm = T)),
      race = first(race),
      .groups = "drop"
    )
)
res <- tidy(mod, effects = "fixed")
means <- data.frame(emmeans(mod, ~ race + reviewyear,
  lmerTest.limit = 1e6,
  pbkrtest.limit = 1e6
))
kable(means)
```

>>>>

```{r libraries}
#| include: false
library(tidyverse)
library(table1)
library(lmerTest)
library(broom.mixed)
library(emmeans)
library(knitr)
```

```{r data cleaning}
#| include: false
# source("/Users/timvigers/GitHub/CF/Kamyron Jordan/Racial and Ethnic Differences in ETI/create_race_eti_analysis_dataset.R")
load("~/Documents/Work/Vigers/CF/Kamyron Jordan/Racial and Ethnic Differences in ETI/Data_Cleaned/analysis_dataset.RData")
```

@@

```{r}
table1(~ race | Hispanicrace, data=demo)
```

@@

```{r}
table1(~race+Hispanicrace|factor(ReviewYear),data = annual)
```

@@

```{r}
table1(~race,data = demo)
```

@@

```{r}
table1(~Gender+WasGenotyped+F508_num|race,data=demo)
```

@@

```{r}
#| cache: true
by_year = encounter %>%
  group_by(eDWID, reviewyear) %>%
  summarise(across(c(race, Gender, WasGenotyped, F508_num, contains("insurance")), first),
    `Age at Last Visit` = last(na.omit(encounterage)),
    `Mean ppFEV` = mean(GLI_FEV1_pct_predicted, na.rm = T),
    `Weight at Last Visit` = last(na.omit(weight)),
    `Weight %ile at Last Visit` = last(na.omit(weightpercentile)),
    `BMI at Last Visit` = last(na.omit(bmivalue)),
    `BMI %ile at Last Visit` = last(na.omit(bmipercentile)),
    across(contains("Vx"), ~ max(.x)),
    .groups = "drop"
  ) %>%
  select(
    -eDWID, reviewyear, race, Gender, WasGenotyped, F508_num, contains("insurance"),
    `Age at Last Visit`, `Mean ppFEV`, `Weight at Last Visit`,
    `Weight %ile at Last Visit`, `BMI at Last Visit`,
    `BMI %ile at Last Visit`, contains("Vx")
  ) %>%
  rename(any_of(tidy_labels)) %>%
  tbl_summary(
    by = Year, missing_text = "Missing",
    statistic = list(all_continuous() ~ "{mean} ({sd})")
  ) %>%
  bold_labels() %>%
  as_kable_extra()
```

# ETI eligibility

Trikafta originally approved October 21, 2019 for patients 12+ years of age. Approved for ages 6 - 11 on June 9, 2021.

```{r}
encounter %>%
  mutate(eti_elig = factor(eti_elig, ordered = T)) %>%
  group_by(eDWID, reviewyear) %>%
  summarise(
    eti_elig = max(eti_elig), race = first(na.omit(race)),
    .groups = "drop"
  ) %>%
  select(-eDWID) %>%
  rename(any_of(tidy_labels)) %>%
  tbl_summary(by = `Race/Ethnicity`, missing = "Missing") %>%
        bold_labels() %>%
        as_kable_extra()
  tbl_strata(
    strata = Year,
    .tbl_fun =
      ~ .x %>%

  )
```

@@

```{r}
#| warning: false
mod_df
pivot_longer(c("GLI_FEV1_pct_predicted", "GLI_FVC_pct_predicted")) %>%
  ggplot(aes(x = reviewyear, y = value, color = factor(eDWID), group = eDWID)) +
  geom_line(alpha = 0.01) +
  theme_bw() +
  theme(legend.position = "none") +
  facet_wrap(~name)
```

```{r}
#| cache: true
# Most basic model of lung function over time by race
mod <- lmer(
  GLI_FEV1_pct_predicted ~ factor(reviewyear) * race + (1 | eDWID),
  data = mod_df
)
res <- tidy(mod, effects = "fixed")
means <- data.frame(emmeans(mod, ~ race + reviewyear,
  lmerTest.limit = 1e6,
  pbkrtest.limit = 1e6
))
kable(means)
```
